# (APPENDIX) Appendix {-}
\cleardoublepage
\phantomsection
\addcontentsline{toc}{part}{Appendices}
<!-- \appendixpage* -->
<!-- \setlength\beforechapskip{-\baselineskip} -->

# Supplement to Chapter 4

```{r include=FALSE}
def_counter <- 0

inc <- function(x) {
  eval.parent(substitute(x <- x + 1))
  x
}
```

## Definitions `r inc(def_counter)` — Graphs, nodes, links

We recall that a graph is formally defined as a pair $G=(N, L)$. Where $N$ is a set whose elements are the nodes (also called vertices or points), and $L$ is a set of links (also called edges) which are ordered pairs of distinct nodes. $L$ is a subset of the set of all possible links between nodes, where in our case a node cannot be associated to itself: $L \subseteq\left\{(x, y)|(x, y) \in N^{2} \wedge x \neq y\right\}$. For a link $(x, y)$, $x$ and $y$ are called the _endpoints_ of a link. In our approach, links are _undirected_ and a link represents an occurrence of $x$ in $y$. As a shorthand we write the link between $x$ and $y$ as $l_{xy}$. For example, the link $l_{xy}$ can represent the occurrence of a category 'date of birth' (node $x$) in the document group 'Eurodac' (node $y$). All categories present in the document group have corresponding nodes and links in the graph. In fact, we can treat each data model (i.e., document group) as a separate graph. The complete graph is then in effect the combination of different data models. For each data model as a separate graph $G_{i}$, the combined graph is the disjoint union of graphs: $G=\bigcup_{i \in I} G_{i}$.

## Definitions `r inc(def_counter)` — Attributes

In our graph model, nodes are objects composed of attributes that are used to keep metadata of nodes. These attributes are formulated using the notation $n.a$ for an attribute $a$ of a node $n$. The most important metadata kept for a node are $n.name$ and $n.type$, where $name$ is the natural language label of the node. The attribute $type$ can only take a limited set of values: $type \in \{category, categoryValue, codeGroup, document, documentGroup\}$.

## Definitions `r inc(def_counter)` — Graph drawing

A drawing of a graph $G=(N,L)$ is a collection of points in a two-dimensional space. Each point $p_i$ with coordinates $x$ and $y$ is the position of the node $n_i$ in the layout. Whenever there exists a link $(p_i, p_j) \in L$, a line is drawn between points $p_i$ and $p_j$. The task of the layout algorithm is to find a positioning of points so that specific criteria are optimally met. Examples of commonly used criteria are: nodes should not overlap, neighbouring nodes should be grouped together, the number of crossing link should be minimised. Each algorithm and set of criteria has its own benefits and drawbacks.

## Definitions `r inc(def_counter)` — Degree & neighbourhood

For a node $n_j$n the degree is defined as the number of links a node has: $deg(x) = \left|\{ n_{j}: l_{ij} \in L \right\}|$. The set of linked nodes is called the _neighbourhood_ of a node. The neighbourhood $H_{i}$ for a node $n_{j}$ is defined as: $H_{i}=\left\{n_{j}: l_{i j} \in L \vee l_{j i} \in L\right\}$.

## Definitions `r inc(def_counter)` — Betweenness centrality

The betweenness centrality of a node $n$ is defined as $bc(n)=\sum_{s \neq n \neq t} \frac{\sigma_{s t}(n)}{\sigma_{s t}}$. Where $\sigma_{st}$ is the total amount of shortest paths from node $s$ to node $t$ and $\sigma_{st}(n)$ is the amount of those paths that pass through $n$. A path is a sequence of nodes, where each pair of nodes in the sequence is linked. The shortest path is the path between two nodes $s$ and $t$ that traverses the smallest number nodes. The equation for betweenness centrality takes into account that there may be several possible paths from $s$ to $t$, with only some passing through $n$.

## Definitions `r inc(def_counter)` — Presence

The presence of all categories in a document group node $n_x$ is a set of all category nodes $Categories(x) = \left\{ n_y \in N: (l_{xy} \in L \vee l_{yx} \in L) \wedge n_y.type=category\right\}$. The presence of a category $n_x$ in a document group is the set of nodes of the type document group for which there exist a link between this category and the document group. Formally defined as: $Presence(x, documentGroup)=\left\{ n_y \in N : l_{xy} \in Links \wedge y.type = docGroup\right\}$.

## Definitions `r inc(def_counter)` — Intersection and difference

The absence of categories between a $docGroup_1$ and $docGroup_2$ is the set of categories present in the second document group minus the set of categories present in the first. In our notation: $Absence(docGroup_1, docGroup_2) = \left\{ Categories(docGroup_2) \setminus Categories(docGroup_1)\right\}$. The categories that are common between those same two document groups are determined using the intersection of the sets of categories that are present in either: $CommonCodes(docGroup_1, docGroup_2)=docGroup_1 \cap docGroup_2$. This operation is not limited to two sets. The intersection between more sets can be notated as $\bigcap_{i=1}^n Presence(docGroup_i)$.

## Table presence of code groups for authorities

```{r code = readLines("./analysis/presence_matrix_authorities.R"), echo = FALSE}
```

```{r echo=FALSE}
capTabAuthorities <- ifelse(usePlaceholderFigures, placeholderCaption, "Presence of code groups for EU (Eurodac, SIS, VIS), Greek (HRF), German (GRF), and their relative degree and betweenness centrality.")
```

```{r authorities, ft.align="center", tab.cap=capTabAuthorities, echo = FALSE}
if (usePlaceholderFigures) {
  placeholder <- c("")
  df <- data.frame(placeholder)
  ft <- flextable(df)
  ft
} else {
  ft <- width(ft, j = 1, width = 0.7)
  ft <- width(ft, j = 2, width = 1)
  ft <- width(ft, j = 3, width = 0.5)
  ft <- align_text_col(ft, align = "left", header = TRUE, footer = TRUE)
  ft <- align_nottext_col(ft, align = "left", header = TRUE, footer = TRUE)
  ft <- theme_booktabs(ft)
  ft
}
```
